---
title: CTmax acclimation in Diacyclops navus
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
---

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/markdown/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}

theme_matt_facets = function(base_size = 18,
                             dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  theme_bw(base_family="sans") %+replace% 
    theme(
      panel.grid = element_blank(),
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      strip.text.x = element_text(size = base_size),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

## Project Description

This project tests how quickly *Diacyclops navus* acclimates to increases and decreases in temperature. Individuals were taken from a bulk culture, maintained in lab at 16°C with a 12:12 ligh:dark cycle and fed ad libitum *Cryptomonas ozolini*, cultured semi-continuosly in WC medium. 

During Summer 2025 we ran `r max(ctmax_data$exp_rep)` experimental replicates. For each experiment, groups of females (6 replicates per experiment, 15 individuals per replicate) were collected and placed into 50 mL of dilute instant ocean (1g / L). These replicates were fed, and allowed to acclimate to the new conditions for one day at 16°C, using a benchtop incubator. After 1 day, CTmax was measured for randomly selected individuals from each replicate (n = 11 individuals per day). After this baseline CTmax measurement, replicates were divided into three groups (n = 2 replicates per group), which were distributed to either a control (16°C), warming (22°C), or cooling (10°C) treatments. CTmax was then measured daily for several indviduals from all three treatments. 

## CTmax Measurements

Total sample sizes (adding across experimental replicates) are shown below. 

```{r}

ctmax_data %>% 
  group_by(day, acc_temp) %>% 
  count() %>% 
  pivot_wider(names_from = acc_temp, values_from = n) %>% 
  janitor::clean_names() %>% 
  select("Day" = day, "10°C" = x10, "16°C" = x16, "22°C" = x22) %>% 
  knitr::kable()

```

CTmax data, again combined across all replicates is shown below, plotted against acclimation duration. 

```{r}
ctmax_data %>% group_by(day, acc_temp) %>% 
  summarise(mean_ctmax = mean(ctmax, na.rm = T)) %>% 
  ungroup() %>% 
  ggplot(aes(x = day, y = mean_ctmax, colour = factor(acc_temp))) + 
  geom_point(data = ctmax_data, aes(y = ctmax), 
             alpha = 0.5, 
             position = position_jitter(width = 0.05, height = 0)) +
  geom_line(linewidth = 1.5) + 
  geom_point(size = 3) + 
  scale_colour_manual(values = c("slateblue", "forestgreen", "gold")) + 
  labs(colour = "Acc. Temp.") +
  theme_matt() + 
  theme(legend.position = "right")
```

Data here is shown for each experimental replicate separately. 

```{r}
ctmax_data %>% group_by(exp_rep, day, rep_id, acc_temp) %>% 
  summarise(mean_ctmax = mean(ctmax, na.rm = T)) %>% 
  ungroup() %>% 
  ggplot(aes(x = day, y = mean_ctmax, colour = factor(acc_temp), group = rep_id)) + 
  facet_wrap(exp_rep~.) + 
  geom_point(data = ctmax_data, aes(y = ctmax), 
             alpha = 0.5, 
             position = position_jitter(width = 0.05, height = 0)) +
  geom_line(linewidth = 1.5) + 
  geom_point(size = 3) + 
  scale_colour_manual(values = c("slateblue", "forestgreen", "gold")) + 
  labs(colour = "Acc. Temp.") +
  theme_matt() + 
  theme(legend.position = "right")
```

## Model

To examine the effect of acclimation over time, we fit a linear mixed effects model to the data. CTmax was modeled as a function of day and acclimation temperature (with interaction). The replicate, tube position during the CTmax assay, and the experimental replicate were all included as random effects. Model performance was checked against several sets of assumptions. 

```{r}
model_data = ctmax_data %>% 
  mutate(day = factor(day), 
         acc_temp = factor(acc_temp))

acc.model = lme4::lmer(ctmax ~ day * acc_temp + (1|rep) + (1|tube) + (1|exp_rep), 
                       data = model_data)

performance::check_model(acc.model)
```

The model indicates a significant interaction between acclimation temperature and the acclimation duration. 

```{r}
car::Anova(acc.model, type = "III") %>% 
  knitr::kable()
```

Marginal means for each acclimation temperature x duration combination were extracted from this model. Contrasts were generated using a 'treatment vs. control' approach, where the 16°C acclimation treatment was designated as the control. These contrasts (treatment CTmax - control CTmax) are shown below.  

```{r}
emmeans::emmeans(acc.model, "acc_temp", by = "day") %>% emmeans::contrast("trt.vs.ctrlk", ref = 2) %>% 
  plot() +  
  geom_vline(xintercept = 0) + 
  labs(x = "Change in CTmax Relative \nto Control (°C)",
       y = "Treatment") + 
  #coord_flip() + 
  scale_y_discrete(labels = c("16°C Effect", "22°C Effect")) + 
  theme_bw(base_size = 24) + 
  theme(panel.grid = element_blank())

```

The acclimation effect steadily increases with time for the cold acclimation treatment (meaning CTmax gradually decreased over time). Warm acclimation, however, had no apparent effect on CTmax. 

```{r}
emmeans::emmeans(acc.model, "acc_temp", by = "day") %>% emmeans::contrast("trt.vs.ctrlk", ref = 2) %>% data.frame() %>% 
  mutate(contrast = if_else(contrast == "acc_temp10 - acc_temp16", "Cold Acc.", "Hot Acc."),
         day = as.numeric(as.character(day))) %>% 
  ggplot(aes(x = day, y = estimate, colour = contrast)) + 
  geom_hline(yintercept = 0) +
  geom_point(size = 3) + 
  geom_smooth(method = "lm", linewidth = 2) + 
  scale_colour_manual(values = c("Cold Acc." = "dodgerblue", 
                                 "Hot Acc." = "indianred")) + 
  labs(x = "Acclimation Duration",
       y = "Acclimation Effect (°C)") + 
  theme_matt() + 
  theme(legend.position = "right")
```


## Acclimation Response Ratios

Examining the relationship between CTmax and acclimation temperature is traditionally how acclimation is measured. The magnitude of acclimation is often standardized against the difference in acclimation temperatures, yielding Acclimation Response Ratios (ARRs; the °C change in thermal limits per °C change in acclimation temperature). 

The relationship between CTmax and acclimation temperature is shown here, broken down by day. A loess smoother is included in each panel to show the relationship between CTmax and acclimation temperature. Notably, reaction norms are relatively flat on Day 1, begin to take on a linear trend in Days 2 and 3, but by Day 6 there is a distinct non-linear trend. 

```{r}

ctmax_data %>% 
  filter(ctmax > 31) %>% 
  # filter(day != 0) %>% 
  # group_by(exp_rep, day, acc_temp) %>% 
  # summarise(mean_ctmax = mean(ctmax, na.rm = T)) %>% 
  # ungroup() %>% 
  # mutate(acc_temp = as.numeric(as.character(acc_temp)),
  #        day = as.numeric(as.character(day))) %>% 
  ggplot(aes(x = acc_temp, y = ctmax)) + 
  facet_wrap(day~.) + 
  geom_point(size = 3, position = position_jitter(width = 0.3, height = 0),
             alpha = 0.5) + 
  geom_smooth(se = F, colour = "black", linewidth = 2) + 
  scale_x_continuous(breaks = c(10, 16, 22)) + 
  theme_matt() + 
  theme(legend.position = "right")

```

As a consequence of these changes, the ARR changed over time, as shown below. 

```{r}

daily_arr = ctmax_data %>% 
  split(ctmax_data$day) %>% 
  map(\(df) lm(ctmax ~ acc_temp, data = df)) %>% 
  map(coef) %>% 
  map_dbl("acc_temp") %>% 
  data.frame() %>% 
  rownames_to_column(var = "day") %>% 
  janitor::clean_names() %>% 
  select(day, "arr" = x) %>% 
  mutate(day = as.numeric(as.character(day))) 


ggplot(daily_arr, aes(x = day, y = arr)) + 
  geom_hline(yintercept = 0) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm", formula = y~log(x + 0.06)) + 
  theme_matt()

```

The same analysis was performed with just the cold acclimation and control treatments. 

```{r}

subset_daily_arr = ctmax_data %>% 
  filter(acc_temp != 22) %>% 
  split(filter(ctmax_data, acc_temp != 22)$day) %>% 
  map(\(df) lm(ctmax ~ acc_temp, data = df)) %>% 
  map(coef) %>% 
  map_dbl("acc_temp") %>% 
  data.frame() %>% 
  rownames_to_column(var = "day") %>% 
  janitor::clean_names() %>% 
  select(day, "arr" = x) %>% 
  mutate(day = as.numeric(as.character(day))) 

ggplot(subset_daily_arr, aes(x = day, y = arr)) + 
  geom_hline(yintercept = 0) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm", formula = y~log(x + 0.06)) + 
  theme_matt()

```

The pattern in ARR is similar until Day 6, when the 'cold only' subset results in a substantially larger ARR than the entire data set. 

```{r}

bind_rows(mutate(daily_arr, "set" = "full"), 
          mutate(subset_daily_arr, "set" = "cold only")) %>% 
  mutate(day = factor(day)) %>% 
  ggplot(aes(x = day, y = arr, fill = set)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) + 
  scale_fill_manual(values = c("dodgerblue", "indianred")) + 
  theme_matt()

```

